---
title: "Parsing M-bias output"
author: Peter Hickey
date: 13 April 2014
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(eval = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
The `bismark_methylation_extractor` produces simple plots of M-bias. While useful, I find these plots a little limited and would like more control over the final figures. Furthermore, I would like to be able to automate the process of choosing how much of the start and end of each read should be ignored based on the M-bias results.

# File format
The output file of `bismark_methylation_extractor --mbias_only` (`<sampleName>.M-bias.txt`) isn't a simple tabular file. 

For single end data it looks something like this:
```
CpG context
===========
position        count methylated        count unmethylated      % methylation   coverage
1       3959916 1669238 70.35   5629154
.
.
.
86      831931  421630  66.37   1253561

CHG context
===========
position        count methylated        count unmethylated      % methylation   coverage
1       636981  24827050        2.50    25464031
.
.
.
86      29437   6969134 0.42    6998571

CHH context
===========
position        count methylated        count unmethylated      % methylation   coverage
1       1797478 73925548        2.37    75723026
.
.
.

86      79795   19761957        0.40    19841752
```

For paired end data it looks something like this:
```
CpG context (R1)
================
position        count methylated        count unmethylated      % methylation   coverage
1       2701458 643907  80.75   3345365
.
.
.
101     1536977 432208  78.05   1969185

CHG context (R1)
================
position        count methylated        count unmethylated      % methylation   coverage
1       1240592 13879394        8.20    15119986
.
.
.
101     81480   9630678 0.84    9712158

CHH context (R1)
================
position        count methylated        count unmethylated      % methylation   coverage
1       3415842 44731045        7.09    48146887
.
.
.
101     245788  32786176        0.74    33031964

CpG context (R2)
================
position        count methylated        count unmethylated      % methylation   coverage
1       511797  5938003 7.94    6449800
.
.
.
101     1864777 0       100.00  1864777

CHG context (R2)
================
position        count methylated        count unmethylated      % methylation   coverage
1       907847  13406746        6.34    14314593
.
.
.
101     60106   0       100.00  60106

CHH context (R2)
================
position        count methylated        count unmethylated      % methylation   coverage
1       2837397 44258539        6.02    47095936
.
.
.
101     164300  0       100.00  164300

```

## `read.mbias()`
I wrote a function to parse these files, `read.mbias()`:
```{r}
read.mbias <- function(file){
  x <- read.table(file, sep = '\n', as.is = TRUE)
  header_lines <- sapply(X = x, function(xx){grepl(pattern = 'context', x = xx)})
  y <- vector(mode = "list", length = sum(header_lines))
  names(y) <- x[header_lines]
  z <- sapply(X = x, function(xx){grep(pattern = "context|^=|position", x= xx, invert = TRUE)})
  dz <- diff(z)
  next_context_idx <- c(which(dz > 1), nrow(z)) # nrow(z) to ensure the last row can be found
  colnames_idx <- sapply(X = x, function(xx){grep(pattern = "position", x = xx)})[1]
  colnames <- strsplit(x = x[colnames_idx, ], split = "\t")[[1]]
  for (i in seq_along(next_context_idx)){
    start_idx <- ifelse(i == 1, 4, z[next_context_idx[i - 1]] + 4) # +4 to skip the intermediate 'header' rows
    end_idx <- z[next_context_idx[i]]  
    yy <- x[seq(from = start_idx, to = end_idx, by = 1), ]
    y[[i]] <- do.call("rbind", lapply(yy, function(yyy, colnames){
      tmp <- strsplit(yyy, split = "\t")
      val <- data.frame(as.integer(tmp[[1]][1]), as.integer(tmp[[1]][2]), as.integer(tmp[[1]][3]), as.numeric(tmp[[1]][4]), as.integer(tmp[[1]][5]))
      colnames(val) <- colnames
      return(val)
      }, colnames = colnames))
    }
  y <- do.call("rbind", y)
  y$Context <- strtrim(rownames(y), width = 3)
  
  ## Add R1/R2 annotation if paired-end data
  if (grepl(pattern = "R1", x = rownames(y)[1])){
    y$Read <- ifelse(grepl(pattern = "R1", x = rownames(y)), "R1", "R2")
  } else{
    y$Read <- "SE"
  }
  rownames(y) <- NULL
  return(y)
  }
```

## `plot.mbias`
__TODOs__
* Make this into a method rather than a function.
* Allow options to be passed to `plot.mbias()`.
* Add sample name to title, etc.

```{r}
library(ggplot2)
plot.mbias <- function(mbias, sample_name){
  presentation_theme <- theme(axis.text = element_text(size = 25, colour = "black"), axis.title.x = element_text(size = 30), axis.title.y = element_text(size = 30, angle = 90), strip.text.x = element_text(size = 25, face = "italic"), strip.text.y = element_text(size = 25, angle = 270, face = "italic"), plot.title = element_text(size = 35, face = "bold"), legend.text = element_text(size = 20), legend.title = element_text(size = 25, face = "bold"))
  if (all(mbias$Read == "SE")){
    ggplot(data = mbias, aes_string(x = "position", y = "`% methylation`")) + geom_line(aes(colour = Context), size = 1) + presentation_theme + ylim(c(0, 100))
    } else{
      #ggplot(data = mbias, aes_string(x = "position", y ="`% methylation`")) + geom_line(aes(colour = Context, linetype = Read), size = 1) + presentation_theme
      ggplot(data = mbias, aes_string(x = "position", y ="`% methylation`")) + geom_line(aes(colour = Context), size = 1) + presentation_theme + facet_grid(Read ~ .) + ylim(c(0, 100)) + ggtitle(paste0(sample_name, ": M-bias"))

    }
}
```

## `trim
```{r}
trim <- function(mbias, mad_cutoff = 3){
  l <- expand.grid(unique(mbias$Context), unique(mbias$Read), stringsAsFactors = FALSE)
  val <- vector("list", nrow(l) / 2)
  for (i in 1:nrow(l)){
    m <- mbias[mbias$Context == l[i, 1] & mbias$Read == l[i, 2], 4]
    val[[i]] <- which(abs(m - median(m)) > mad_cutoff * mad(m))
  }
  names(val) <- paste0(l[, 1], l[, 2])
  return(val)
}
```

## __TODO__
* Display results of `trim` in the plot created by `plot.mbias`.

# Examples

Here I read in the m-bias data for single-end (FF) and paired-end (E13BUF) samples.

## FF: Single end data 
```{r}
setwd('/Volumes/hickey/methylation_m-tuples/')
FF <- read.mbias(file = "FF/FF.M-bias.txt")
plot.mbias(FF, "FF")
```

## E13BUF: Paired end data
```{r}
setwd('/Volumes/hickey/methylation_m-tuples/')
E13BUF <- read.mbias(file = 'E13BUF/E13BUF.M-bias.txt')
plot.mbias(E13BUF, "E13BUF")
```

## Adipose data
```{r, results = 'asis'}
setwd('/Volumes/hickey/methylation_m-tuples/')
sample_names <- expand.grid(c('E13', 'E18', 'E23'), c('BUF', 'SA', 'VA', 'VAT'), stringsAsFactors = FALSE)
sample_names <- paste0(sample_names[, 1], sample_names[, 2])
for(sample_name in sample_names){
  file <- paste0(sample_name, '/', sample_name, '.M-bias.txt')
  mbias <- read.mbias(file = file)
  print(plot.mbias(mbias, sample_name = sample_name))
  trim(mbias, mad_cutoff = 4)
  }
```