---
title: "Unexplained subsetting behaviour of classes based on SummarizedExperiment"
author: Peter Hickey
date: 30 April 2014
output: html_document
---

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Unexplained subsetting behaviour of classes based on SummarizedExperiment}
-->

# Motivation

The `[` method is not working as expected for my simple class that extends the `SummarizedExperiment` class.

# Example
This example shows the unexpected (to me) behaviour.

## `MyClass` definition
Define `MyClass`, basically a `SummarizedExperiment` with an `extraSlot`, which is a numeric vector of the same length as the number of rows in the `rowData`:

```{r, eval = FALSE}
require(GenomicRanges)
.MyClass <- setClass("MyClass", representation(extraSlot = "vector"), contains = "SummarizedExperiment")
```
## `MyClass` constructor 
```{r, eval = FALSE}
MyClass <- function(assays, gr, extraSlot){
  se <- SummarizedExperiment(assays = assays, rowData = gr)
  mc <- .MyClass(se, extraSlot = extraSlot)
  mc
}
```

## `[` method definition for `MyClass`
Define the `[` method for `MyClass` objects. Should basically do the same thing as `[` for a `SummarizedExperiment` but also subset the `extraSlot`:

```{r, eval = FALSE}
setMethod("[", c("MyClass", "ANY", "missing"),
          function(x, i, j, ..., drop = FALSE)
          {
            if (missing(i) && missing(j)){
              return(x)
            }
            if (missing(i)){
              i <- seq_len(nrow(x))
            }
            if (missing(j)){
              j <- seq_len(ncol(x))
            }
            initialize(x, as(x, "SummarizedExperiment")[i, j, ..., drop = drop], extraSlot = x@extraSlot[i])
            })
```

## Example data
```{r, eval = FALSE}
nrows <- 20
ncols <- 6
counts <- matrix(runif(nrows * ncols, 1, 1e4), nrows)
colnames(counts) <- letters[1:6]
gr <- GRanges('chr1', IRanges(floor(runif(nrows, 1e5, 1e6)), width=100))
extraSlot <- runif(n = length(gr))

ex <- MyClass(assays = SimpleList(counts), gr = gr, extraSlot = extraSlot)
ex
slotNames(ex)
ex@extraSlot
```

## Testing the `[` method for `MyClass` objects
As expected, the following contains 10 elements:
```{r, eval = FALSE}
ex[1:10, ]@extraSlot
```
As unexpected, the following contains 20 elements. __Why__?
```{r, eval = FALSE}
ex[1:10, 2]@extraSlot
```
As unexpected, the following contains 10 elements. __Why? This is the exact same code as the final line of the `[` method definition:__
```{r, eval = FALSE}
initialize(ex, as(ex, "SummarizedExperiment")[1:10, 2, drop = FALSE], extraSlot = ex@extraSlot[1:10])@extraSlot
```

# Summary
I never got to the bottom of this. I didn't post a question to the Bioconductor mailing lists. `MyClass` was a simplified version of an earlier incarnation the `CoMeth` class. When I changed the definition of the `CoMeth` class[^cometh_class_definition], I got the intended subsetting behaviour of the `[` method "for free" via inheritance. Therefore, this was no longer a problem for the `cometh` package, although it would be good to understand the behaviour in the example.

[^cometh_class_definition]: The `CoMeth` class definition changed to use the `MTuples` class, which extends the `GRanges` class, as the `rowData`, rather than the `GRanges` class itself. Effectively, the `extraPos` information is stored in the `MTuples` object and preserved in subsetting by use of the `GenomicRanges:::extraColumnSlotNames` function (see class definition of `MTuples` in `R/AllClasses.R` and the vignette _Creating MTuples and CoMeth classes_).

# Session info
```{r, eval = FALSE}
sessionInfo()
```